---
title: "2021 rides"
author: "Antonio"
date: "`r Sys.Date()`"
output: html_document
---

## Setting up my environment
Notes: Setting up my R environment by loading the 'tidyverse',  
'dplyr', 'readxl', and 'lubdridate' packages.

```{r loading packages}
library(tidyverse)
library(dplyr)
library(readxl)
library(lubridate)
```

## Import Our Data
Notes: Here we would be transferring our data from excel into R, the data can be found [ride data](https://divvy-tripdata.s3.amazonaws.com/index.html). What was added into the raw excel data is the difference from end_dration and start_duration and the different days using WEEKDAY function.  

```{r Naming and loading data sheets of each month}
m1_y2021 <- read_excel("~/Downloads/202101-divvy-tripdata(AutoRecovered).xls")
m2_y2021 <- read_excel("~/Downloads/202102-divvy-tripdata.xls")
m3_y2021 <- read_excel("~/Downloads/202103-divvy-tripdata.xls")
m4_y2021 <- read_excel("~/Downloads/202104-divvy-tripdata.xls")
m5_y2021 <- read_excel("~/Downloads/202105-divvy-tripdata.xls")
m6_y2021 <- read_excel("~/Downloads/202106-divvy-tripdata.xls")
m7_y2021 <- read_excel("~/Downloads/202107-divvy-tripdata.xls")
m8_y2021 <- read_excel("~/Downloads/202108-divvy-tripdata.xls")
m9_y2021 <- read_excel("~/Downloads/202109-divvy-tripdata.xls")
m10_y2021 <- read_excel("~/Downloads/202110-divvy-tripdata.xls")
m11_y2021 <- read_excel("~/Downloads/202111-divvy-tripdata.xls")
m12_y2021 <- read_excel("~/Downloads/202112-divvy-tripdata.xls")
```

## Inspect Data
Notes: Here we would be searching for incongruencies
```{r Checking for any errors in column names}
str(m1_y2021)
str(m2_y2021)
str(m3_y2021)
str(m4_y2021)
str(m5_y2021)
str(m6_y2021)
str(m7_y2021)
str(m8_y2021)
str(m9_y2021)
str(m10_y2021)
str(m11_y2021)
str(m12_y2021)
```

## Join Data Frames
Notes: Here we combine all data frames into one single data frame

```{r combining all dataframes into one}
all_2021 <- bind_rows(m1_y2021, m2_y2021, m3_y2021, m4_y2021, m5_y2021, m6_y2021, 
                      m7_y2021, m8_y2021, m9_y2021, m10_y2021, m11_y2021, m12_y2021)
```

## Choosing What Columns to Keep
Notes: Here we remove columns start_station_name, start_station_id, end_station_name, end_station_id, start_lat, start_lng, end_lat, end_lng, sum_casu_class_sun, sum_casu_elec_sun, average, mean_of_ridelength, ...19, ...20, ...21,max_ridelength, most_times_used that we were not going to use

```{r selecting data that is not needed}
all_2021 <- all_2021 %>%  
  select(-c(start_station_name, start_station_id, end_station_name, end_station_id, 
            start_lat, start_lng, end_lat, end_lng, sum_casu_class_sun, 
            sum_casu_elec_sun, average, mean_of_ridelength, ...19, ...20, ...21,
            max_ridelength, most_times_used))
```

## Formatting Date
Notes: Here we added more columns where we can see date, date, month, day, and year of each ride, in order to aggregate the data at the level of the rider. 

```{r adding more columns for date, month, day, year, and day of week}
all_2021$date <- as.Date(all_2021$started_at)
all_2021$month <- format(as.Date(all_2021$date), "%m")
all_2021$day <- format(as.Date(all_2021$date), "%d")
all_2021$year <- format(as.Date(all_2021$date), "%Y")
all_2021$day_of_week <- format(as.Date(all_2021$date), "%A")
```

## Remove poor data
Notes: Removing negative numbers because they dont represent user usage. 

```{r Removing bad data}
all_2021_v2 <- all_2021[!(all_2021$ride_length<0),]
```

## Descriptive analysis
Notes: calculating the overall mean, median, max, and min

```{r overall mean, median, mode}
summary(all_2021_v2$ride_length)
```

## Compared members and casual riders
Notes: Comparing members and casual riders by ride length using the mean, median, max, and min

```{r comparing members and casual}
aggregate(all_2021_v2$ride_length ~ all_2021_v2$member_casual, FUN = mean)
aggregate(all_2021_v2$ride_length ~ all_2021_v2$member_casual, FUN = median)
aggregate(all_2021_v2$ride_length ~ all_2021_v2$member_casual, FUN = max)
aggregate(all_2021_v2$ride_length ~ all_2021_v2$member_casual, FUN = min)
```

## Average time by each day
Notes: Here we calculate the average by each day and rideable type

```{r Comparing average by each day and rideabletype per member and casual riders}
aggregate(all_2021_v2$ride_length ~ all_2021_v2$member_casual + 
            all_2021_v2$day_of_week + all_2021_v2$rideable_type, FUN = mean)
```

## Fixing the order of the days
Notes: Changing the order so that the days go in order

```{r ordering the days}
all_2021_v2$day_of_week <- ordered(all_2021_v2$day_of_week, 
                                    levels=c("Sunday", "Monday", "Tuesday", 
                                             "Wednesday", "Thursday", "Friday", 
                                             "Saturday"))
```

## Making sure the days are in order

```{r checking if days are in order}
aggregate(all_2021_v2$ride_length ~ all_2021_v2$member_casual + 
            all_2021_v2$day_of_week + all_2021_v2$rideable_type, FUN = mean)
```

## Visualizing rides by rider type
Notes: Here we perform some calculations and group by functions, and mutating to create weekday, while creating a visual of the data

```{r Weekday vs. Number of Rides}
all_2021_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%  
  summarise(number_of_rides = n() 
            ,average_duration = mean(ride_length)) %>% 		
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(x = "Weekday", y = "Number of Rides", title = "Weekday vs. Number of Rides", fill = "Member/Casual")
```

## Visualizing Rides by Average Duration
Notes: Here we used the same code as above and this time replaced number of rides to average duration, thus plotting
weekday against average duration.

```{r Weekday vs. Average Duration}
all_2021_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(x = "Weekday", y = "Average Duration", title = "Weekday vs. Average Duration", fill = "Member/Casual")
```

## Visualizing type of bike by number of rides
Notes: Here we perform the same calculations and plot type of bike against number of rides.

```{r bike type vs. average duration}
all_2021_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday, rideable_type) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = rideable_type, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(x = "Bike Type", y = "Average Duration", title = "Bike Type vs. Average Duration", fill = "Member/Casual")
```

## Visualizing type of bike by average time spent
Notes: Here we perform the same calculation but this time plot type of bike against average duration

```{r Bike Type vs. Number or Rides}
all_2021_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday, rideable_type) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x = rideable_type, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(x = "Bike Type", y = "Number of Rides", title = "Bike Type vs. Number of Rides", fill = "Member/Casual")

```

